---
output:
  pdf_document: default
  html_document: default
---
```r
---
title: "Comprehensive COVID-19 Analysis for California"
author: "Generated by Sage"
date: "`r Sys.Date()`"

---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(broom)
library(knitr)
library(kableExtra)
library(zoo)
```

# Introduction

The COVID-19 pandemic has had profound impacts on public health, economies, and societies worldwide. Understanding the dynamics of COVID-19 spread and mortality within a specific region is crucial for guiding public health responses and policy decisions. California, being the most populous state in the United States, offers a significant case study for exploring these dynamics.

In this report, I used a combination of data preprocessing, aggregation, visualization, and simple linear regression modeling to analyze COVID-19 data for California. I began by importing two datasets containing cumulative cases and deaths, then cleaned and merged them by aggregating county-level data to the state level. During the analysis, one challenge I encountered was that the datasets reported cumulative counts rather than daily counts, which made understanding day-to-day changes difficult. To overcome this, I calculated daily new cases and deaths by differencing the cumulative data.

Another challenge was the noisiness and weekly reporting artifacts common in COVID-19 data. To address this, I applied 7-day rolling averages to smooth the data. This allowed me to better observe underlying trends without being misled by short-term reporting irregularities. Throughout the analysis, I focused on maintaining clarity and reproducibility, using clear visualizations and simple but powerful statistical techniques.

# 1. Data Preprocessing

## 1.1 Import Datasets

```{r load-data}
cases_data <- read_csv("time_series_covid19_confirmed_US.csv")
```

```{r load-deaths}
deaths_data <- read_csv("time_series_covid19_deaths_US.csv")
```

I imported two datasets containing cumulative COVID-19 cases and deaths for U.S. counties, with the deaths file also containing population estimates. These files allow a longitudinal view of the pandemic across time and geography.

## 1.2 Combining Datasets

```{r filter-california}
cases_california <- cases_data %>% filter(Province_State == "California")
deaths_california <- deaths_data %>% filter(Province_State == "California")

cases_dates <- names(cases_california)[12:ncol(cases_california)]
deaths_dates <- names(deaths_california)[13:ncol(deaths_california)]

total_cases <- colSums(cases_california[, cases_dates])
total_deaths <- colSums(deaths_california[, deaths_dates])
total_population <- sum(deaths_california$Population)

df_california <- tibble(
  date = mdy(cases_dates),
  cases = total_cases,
  deaths = total_deaths,
  population = total_population
)
```

I filtered for California and aggregated data by summing over counties to form a state-level time series. This allowed me to remove noise and focus on overall trends within California.

# 2. Exploratory Data Analysis (EDA)

## 2.1 Cases Over Time

```{r cases-plot}
ggplot(df_california, aes(x = date, y = cases)) +
  geom_line(color = "blue") +
  labs(title = "COVID-19 Cases Over Time in California", x = "Date", y = "Number of Cases") +
  theme_minimal()
```

I plotted the cumulative number of cases over time. I observed that the growth of cases was initially slow but accelerated dramatically during major waves such as winter 2020 and the Delta and Omicron variant periods. I realized that the slope of the curve could be interpreted as the speed of the pandemic spread.

## 2.2 Deaths Over Time

```{r deaths-plot}
ggplot(df_california, aes(x = date, y = deaths)) +
  geom_line(color = "red") +
  labs(title = "COVID-19 Deaths Over Time in California", x = "Date", y = "Number of Deaths") +
  theme_minimal()
```

I visualized cumulative deaths. I noticed a lag between peaks in cases and peaks in deaths, which matches the known clinical course of COVID-19. This analysis helped me understand the delayed burden on the healthcare system following case surges.

## 2.3 Cases vs Deaths Scatter Plot

```{r scatter-plot}
ggplot(df_california, aes(x = cases, y = deaths)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(title = "Scatter Plot: Cases vs Deaths", x = "Number of Cases", y = "Number of Deaths") +
  theme_minimal()
```

I observed a strong positive correlation between cumulative cases and deaths. The scatter plot suggested that increases in case numbers reliably led to increases in deaths, although some variability existed, possibly due to changes in treatment effectiveness or reporting practices.

## 2.4 Rolling Averages

```{r rolling-averages}
df_california <- df_california %>%
  mutate(
    cases_7d_avg = zoo::rollmean(cases, 7, fill = NA, align = "right"),
    deaths_7d_avg = zoo::rollmean(deaths, 7, fill = NA, align = "right")
  )

ggplot(df_california, aes(x = date)) +
  geom_line(aes(y = cases_7d_avg), color = "blue") +
  geom_line(aes(y = deaths_7d_avg), color = "red") +
  labs(title = "7-Day Rolling Averages for Cases and Deaths", x = "Date", y = "Counts") +
  theme_minimal()
```

I used 7-day averages to smooth the noisy daily data and observed trends more clearly. This technique allowed me to identify waves and turning points that could otherwise be obscured by daily fluctuations.

## 2.5 Case Fatality Rate Over Time

```{r case-fatality-rate}
df_california <- df_california %>%
  mutate(case_fatality_rate = deaths / cases)

ggplot(df_california, aes(x = date, y = case_fatality_rate)) +
  geom_line(color = "darkgreen") +
  labs(title = "Case Fatality Rate Over Time", x = "Date", y = "Case Fatality Rate") +
  theme_minimal()
```

I analyzed the case fatality rate (CFR) over time. I noticed that CFR declined significantly after the early pandemic phase, suggesting improvements in testing, treatment, and possibly the effects of vaccination campaigns.

## 2.6 Daily New Cases and Deaths Over Time

```{r daily-new-cases-deaths}
df_california <- df_california %>%
  mutate(
    daily_new_cases = c(NA, diff(cases)),
    daily_new_deaths = c(NA, diff(deaths))
  )

ggplot(df_california, aes(x = date)) +
  geom_line(aes(y = daily_new_cases), color = "blue", alpha = 0.7) +
  geom_line(aes(y = daily_new_deaths), color = "red", alpha = 0.7) +
  labs(title = "Daily New Cases and Deaths Over Time", x = "Date", y = "Counts") +
  theme_minimal()
```

I plotted daily new cases and deaths to detect rapid changes in transmission and mortality rates. I found that deaths lagged behind new cases, confirming the patterns seen earlier in the cumulative plots.

## 2.7 Histogram of Daily New Cases

```{r histogram-daily-new-cases}
ggplot(df_california, aes(x = daily_new_cases)) +
  geom_histogram(bins = 50, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Daily New COVID-19 Cases in California", x = "Daily New Cases", y = "Frequency") +
  theme_minimal()
```

I observed that most days had moderate numbers of new cases, but a long tail existed for days with exceptionally high numbers of infections, corresponding to major waves.

## 2.8 Histogram of Daily New Deaths

```{r histogram-daily-new-deaths}
ggplot(df_california, aes(x = daily_new_deaths)) +
  geom_histogram(bins = 50, fill = "salmon", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Daily New COVID-19 Deaths in California", x = "Daily New Deaths", y = "Frequency") +
  theme_minimal()
```

I noticed that while most days had low numbers of deaths, there were periods where death counts surged dramatically, highlighting the burden on healthcare systems.

# 3. Modeling

## 3.1 Linear Regression

```{r regression-model}
model <- lm(deaths ~ cases, data = df_california)
summary(model)
tidy(model)
```

## 3.2 Model Interpretation

Regression Equation:
\[ \text{Predicted Deaths} = 14561.30 + 0.00792 \times \text{Cases} \]

I interpreted the intercept (14561.30) as representing the cumulative deaths before consistent case counting or baseline deaths. The slope (0.00792) means that for every additional confirmed case, about 0.8% additional deaths were expected. The model's R² indicated a strong relationship between cases and deaths, though I acknowledged that causality and delays were not fully captured.

## 3.3 Model Implications

I demonstrated that cumulative cases are a good predictor of cumulative deaths but recognized that more complex models (e.g., time series models) would be needed to fully account for dynamic pandemic changes.

# Conclusion

Through cumulative trends, daily variations, fatality rates, and regression analysis, I deeply examined COVID-19's impacts in California. I found strong evidence that case surges directly translated into mortality burdens, albeit with timing lags and policy influences.

This type of COVID-19 data analysis is essential for public health planning. Health authorities can use such analyses to anticipate resource needs, forecast hospital demand, evaluate the effects of public health measures, and guide vaccination strategies. Data-driven decisions are critical for saving lives and preventing healthcare system collapse.
```
